---
title: "Titanic_survival_analysis_report"
author: "YutongLIU"
date: "2016年12月19日"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is the first time I tried to do a data analysis practice by myself after data science course. Any comments and recommendation about this practice is welcome :)

Here is the agenda of this report

> 1 Introduction
> 2 Defining the Question
  + 2.1 What is the outcome we want?
  + 2.2 What data can we use to answer the question?
> 3 Feature Engineering
  + 3.1 Missing data imputation
  + 3.2 Which varialbes may be relevant to the outcome?
    * 3.2.1 Higher Passenger class, more possible to survive?
    * 3.2.2 Children and women first?
    * 3.2.3 Families can help each other?
    * 3.2.4 Is cabin location relevant to the passenger survival?
> 4 Prediction
> 5 Conclusion
----------------------------------------------------------------

> 1 Introduction

With this dataset, let's find out if it's possible to predict Titanic passengers' survival based on their profile.

Firstly I downloaded the dataset here [link](https://www.kaggle.com/c/titanic)

### Set working directory
```{r preparation,results = 'hide'}
# Set working directory
if(!dir.exists("D:/kaggle/kaggle_Titanic"))
        dir.create("D:/kaggle/kaggle_Titanic")
setwd("D:/kaggle/kaggle_Titanic")

# Load packages
library(dplyr)   # cleaning dataset
library(ggplot2) # drawing graphics for exploratory analysis
library(rpart)   # missing data imputation
library(caret)   # prediction
library(randomForest) #prediction
```

### Loading training dataset

Here are the reasons why I save integer variables in different formats
  + **Survived**: I think it's nomial variable, so I save it as factor.
  + **Pclass**  : I think it's ordinal variable, so I save it as integer.
  + **Sibsp** and **Parch**: maybe I will try to calculate some new variables, so keep them as integer.
  
```{r download_read_dataset,results = 'hide'}
# Download and read dataset
colclass <- c("integer","factor","integer","character",
            "factor","numeric","integer","integer",
            "character","numeric","character","factor")
train_raw <- read.csv(file="./train.csv", header = TRUE,
                      colClasses = colclass,
                       na.strings = c("","NA"),stringsAsFactors = FALSE)  

# Do the same process to the test set
# There is no "survived" variable, so change the column class
colclass_test<-c("integer","integer","character",
            "factor","numeric","integer","integer",
            "character","numeric","character","factor")
test_raw <- read.csv(file="./test.csv", header=TRUE, 
                     colClasses=colclass_test,
                      na.strings = c("","NA"),stringsAsFactors = FALSE)
```

Now let's have a glance at the dataset 
```{r data_structure,results = 'hide'}
str(train_raw)
```

There are 11 variables(**PassengerID**　excluded), here is the definition of each variables.

Variable|Description
--------|-----------
survival|Survival(0 = No; 1 = Yes)
pclass  |Passenger Class(1 = 1st; 2 = 2nd; 3 = 3rd)
name    |Passenger name
sex     |Sex
age     |Age
sibsp   |Number of siblings/spouses aboard
parch   |Number of parents/children aboard
ticket  |Ticket number
fare    |Passenger fare
cabin   |Cabin
embarked|Port of Embarkation(C = Cherbourg; Q = Queenstown; S = Southampton)

> 2 Defining the Question
   + 2.1 What is the outcome we want?
Clearly it is important to set the target of data analysis before anything gets start.
We want to predict the survival of Titanic passengers. In the dataset, **pclass** is the outcome we will predict in testset.
   + 2.2 What data can we use to answer the question?
There are 10 variables, two of them, **age** and **sex**, are demographic information, there is also passengers' name. I think another 7 varibales can be categorized to information about their trip: **embarked**, where did they get on the boat; **sibsp** and **parch**, were they alone or accompanied; **pclass**, which class are they, etc.


> 3 Feature Engineering
  + 3.1 Dataset cleaning and Missing data imputation

### 3.1.1 Dataset cleaning
#### Combination
To ensure the preprocess is the same to train and also test set, now we combine them together.
```{r combination,results = 'hide'}
total_raw <- bind_rows(train_raw,test_raw)

# Save dataset in a new data frame
total_clean <- total_raw
```

#### Text Editing  
Firstly, let's check the variable names, are they easy to use? Fortunately there is no space among the variable names. Now I just standardize them in lower case.
```{r text_edit,results = 'hide'}
# Standardize all variable names in lower case
names(total_clean) <- tolower(names(total_raw))
```

### 3.1.2 Missing value inputation
#### a) Counting missing values in each variable
```{r counting_missing_values,results = 'hide'}
na_count<-summary(apply(total_clean,2,is.na))
```

From the table below, we learned that there are NA values in several variables
Because the dataset contains the test set, there should be missing values in **survived** so it won't be shown in the table, and **passengerid** won't be used as predictor, therefore it won't be shown in the table neither.

Variables                         |NA values #
----------------------------------|-----------
'r dimnames(na\_count)[[2]][[3]]' |'r na_count[3,3]'
'r dimnames(na\_count)[[2]][[4]]' |'r na_count[3,4]'
'r dimnames(na\_count)[[2]][[5]]' |'r na_count[3,5]'
'r dimnames(na\_count)[[2]][[6]]' |'r na_count[3,6]'
'r dimnames(na\_count)[[2]][[7]]' |'r na_count[3,7]'
'r dimnames(na\_count)[[2]][[8]]' |'r na_count[3,8]'
'r dimnames(na\_count)[[2]][[9]]' |'r na_count[3,9]'
'r dimnames(na\_count)[[2]][[10]]'|'r na_count[3,10]'
'r dimnames(na\_count)[[2]][[11]]'|'r na_count[3,11]'
'r dimnames(na\_count)[[2]][[12]]'|'r na_count[3,12]'

#### b) Finding observations that contain missing value and process

##### Age
There are 263 NA values in variable **age**.
```{r age_imputation}
set.seed(123)
factor_name<-c("survived","sex")
total_clean[factor_name]<-lapply(total_clean[factor_name], function(x) as.factor(x))
# Age missing value imputation with rpart
fit_age <- rpart(age ~ pclass+sex+sibsp+parch,
               data=total_clean[!is.na(total_clean$age),3:12],
               na.action=na.omit,method = "anova")
age_result <- predict(fit_age,total_clean[,3:12])
age_result <- round(age_result)
imp_age<-total_clean[,6]
imp_age[is.na(imp_age)]<-age_result[is.na(imp_age)]

# plot age distribution
par(mfrow=c(1,2))
hist(total_raw$Age, freq=F, main='Age: Original Data',col='blue', 
     ylim=c(0,0.07))
hist(imp_age$age, freq=F,main='Age: Imputed Data',col='lightblue', 
     ylim=c(0,0.07))

#impute missing value in age
total_clean$age<-imp_age$age
sum(is.na(total_clean$age))
```

##### Fare
After checking the observation with NA in fare, we find that this passenger is third class and embarked from Southampton. Based on the distribution of third class ticket fare from Southampton, we use the median to impute the NA value.
```{r fare_imputation}
# Check the observation with NA in fare
total_clean[is.na(total_clean$fare),1:12]

# distribution of fare
g_fare<-ggplot(data =total_clean[which(total_clean$pclass==3),], 
               aes(embarked,fare))
g_fare+geom_boxplot()
total_clean[is.na(total_clean$fare),"fare"]<-
        median(total_clean[total_clean$pclass=="3"&total_clean$embarked =="S",]$fare,na.rm = TRUE)
```

##### Cabin
It seems that the number of cabin is not related to passenger survival
so if the cabin is missing value, we think the passenger didn't have a cabin
and we change this variable to a logical vector
```{r cabin_missing_value}
#missing value in cabin
total_clean<-mutate(total_clean,iscabin=is.na(cabin))
total_clean$iscabin<-factor(total_clean$iscabin,levels=c(TRUE,FALSE),
                               labels=c("no","yes"))
cabin <- total_clean$cabin
cabin_cap<-gsub("[0-9]","",cabin)
unique(cabin_cap)
cabin_cap<-as.factor(cabin_cap)
cabin_level<-c("C","E","G","D","A","C","B","F","F","D","B","E","C","B","T","B","E")
levels(cabin_cap)<-cabin_level
total_clean$cabin<-cabin_cap

cabin_survived <-  total_clean[1:891,]
g_cabin <- ggplot(data = cabin_survived[!is.na(cabin_survived$cabin),],
                aes(cabin,fill=survived))
g_cabin+geom_bar()
# fare vs cabin boxplot
g_cab_fare <- ggplot(data = cabin_survived[!is.na(cabin_survived$cabin),],
                aes(x=cabin,y=fare))
g_cab_fare+geom_boxplot()
```


##### Embarked
```{r embarked}
total_clean$embarked[c(62, 830)] <- 'C'
```


3. Grab information and create new variables

Address from passenger name
```{r title}
# Variable "Name" contains two types of information: 
# name and also sex&marital status
# Extract sexmarital status

nametitle<-gsub("(.*, )|(\\..*)","",x=total_clean$name)

#As the factor levels in the train dataset may be different from the test set
#So we don't coerse this variable into factor.

#there are some repeated title so we combined those titles.
special_male<-c("Capt","Col","Don","Dr","Jonkheer",
                "Major","Master","Rev","Sir","the Countess")
special_female<-c("Lady","Dona","Mlle","Mme","Ms")
nametitle[nametitle %in% special_male]<- "Special_male"
nametitle[nametitle %in% special_female]<- "Special_female"

total_clean<-mutate(total_clean,title=nametitle)
total_clean$title<-as.factor(total_clean$title)
```

#### Family size

```{r family_size}
#为什么赋值顺序换为single,small,large，到large时就会全部被赋值为large?
family_size<-total_clean$sibsp+total_clean$parch+1
family_size[family_size > 4]<-"large"
family_size[family_size > 1 & family_size < 5]<-"small"
family_size[family_size == 1]<-"single"
total_clean<-mutate(total_clean,familysize=as.factor(family_size))
```
(把姓提取出来，所有parch+sibsp!=0的，同一个姓归为一组，即一个家庭，是否年龄稍小的存活几率大？)
```{r family_group}
fam_group <- total_clean[1:891,c("survived","name","age","sibsp","parch")]
# grab the family name
fam_name<-gsub(",.*\\..*","",x=fam_group$name)
fam_group<-mutate(fam_group,fam_name=fam_name)
# grab the family size
fam_group<-mutate(fam_group,fam_size=parch+sibsp)
# delete those who were on board alone
fam_group<-fam_group[which(fam_group$fam_size!=0),]
# group family
fam_group<-group_by(fam_group,fam_name)
# find the youngest one in a family
fam_group<-mutate(fam_group,role=age-min(age))
fam_group[which(fam_group$role!=0),"role"]<-"elder"
fam_group[which(fam_group$role==0),"role"]<-"youngest"

#yes!
```

```{r family_role}
fam_name<-gsub(",.*\\..*","",x=total_clean$name)
fam_group<-mutate(total_clean,fam_name=fam_name)
fam_group<-group_by(fam_group,fam_name)
fam_group<-mutate(fam_group,age_var=age-min(age))

fam_group<-mutate(fam_group,role = c("alone"))
fam_group[which(fam_group$parch+fam_group$sibsp!=0 && fam_group$age_var!=0),"role"]<-"elder"
fam_group[which(fam_group$parch+fam_group$sibsp!=0 && fam_group$age_var==0),"role"]<-"youngest"
total_clean<-mutate(total_clean,role=fam_group$role)
```
### Exploratory analysis & data process

### Feature selection
features: pclass,sex,age,fare,iscabin,title,familysize

```{r split_data}
trainset<-total_clean[1:891,c("survived","pclass","sex","age","fare","title","familysize","role")]
testset<-total_clean[892:1309,]
trainset$embarked<-as.factor(trainset$embarked)
testset$embarked<-as.factor(testset$embarked)

```

### Model selection
```{r model}

#test
trainset<-total_clean[1:891,]
trainset$role<-as.factor(trainset$role)
rf1<-randomForest(survived ~ pclass + sex + age + fare +
                          title + familysize + role,
                  data = trainset)
rf2<-randomForest(survived ~ pclass + sex + age + fare +
                          iscabin + title + familysize+
                          sibsp+parch+embarked,
                  data = trainset)
```

### Prediction
```{r prediction}
rf1_result<-predict(rf1,trainset)
confusionMatrix(trainset$survived,rf1_result)

rf2_result<-predict(rf2,trainset)
confusionMatrix(trainset$survived,rf2_result)
```


### Predict with test data
```{r predict_test_data}
# pay attention to the type of every predictor!
testset<-total_clean[892:1309,]
testset$role<-as.factor(testset$role)
levels(testset$role)<-c("alone","youngest")

test6<-predict(rf1,testset)
titanic_rf_submission6<-
        data.frame(Passengerid = testset$passengerid,
                   Survived = test6)
write.csv(x = titanic_rf_submission6,
          file = "./titanic_rf_submission6.csv",row.names = FALSE)
```

